---
title: "EDA"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(gt)
library(ggplot2)
```

## Read in Data

```{r}
nflData = read.csv(file = "data/nflData.csv")
head(nflData)
```

## Select Variables we will work with:

```{r}
names(nflData)

```

```{r}

## Remove any cols with prob:

nflDataClean = nflData |> select(-contains("prob"))

## Select Desired Columns:


nflDataClean = nflDataClean |> select(
  Date,
  qtr,
  posteam,
  ydstogo,
  PassOutcome,
  YardsAfterCatch,
  InterceptionThrown,
  GameID,
  TimeSecs,
  yrdln,
  ydsnet,
  Touchdown,
  PassLength,
  AbsScoreDiff,
  Season,
  Drive,
  time,
  yrdline100,
  Yards.Gained,
  PlayType,
  PassAttempt,
  HomeTeam,
  AwayTeam,
  PosTeamScore,
  DefTeamScore
) |>
  filter(posteam != "")

```

## Cast Variables to Correct Type:

-   QTR

-   Passoutcome

-   HomeTeam

-   AwayTeam

-   PlayType

-   Season

-   GameID

## Cast Categorical Variables to Factors

```{r}

factors = c("qtr",
            "posteam",
            "PassOutcome",
            "HomeTeam",
            "AwayTeam",
            "PlayType",
            "Season",
            "GameID",
            "Touchdown",
            "PassAttempt"
            )
nflDataClean[factors] = lapply(nflDataClean[factors], as.factor)


head(nflDataClean)
```

## Summarize Numeric Variables:

I decided to summaries across team and season, will use this for later

```{r}

summaries = nflDataClean |> group_by(posteam) |>
  summarize(across(where(is.numeric), 
                   list("mean" = mean), 
                   .names = "{.fn}_{.col}"))

gt(summaries)

```

```{r}


```

## Contingency Tables

### Types of Plays

```{r}

plays = nflData |> group_by(PlayType)|> select(PlayType) |> summarise(count = n()) |> arrange(desc(count))

gt(plays) |> tab_header(title = "Types of Plays Run")

nflData |> filter(PlayType == "End of Game") |> select(PosTeamScore)


```

## Pass Length vs Pass Result

```{r}
passResults = nflData|> select(PassOutcome, PassLength) |> filter(PassLength == "Deep" | PassLength == "Short") |> group_by(PassOutcome, PassLength) |> summarize(count = n()) |> pivot_wider(names_from = PassOutcome, values_from = count)



gt(passResults) |> tab_header("Outcome of Pass vs Pass Length")
```

## Pass vs Run Contingency Table:

```{r}

passAndRunAttempts = nflDataClean |> 
  filter(PlayType == "Run" | PlayType == "Pass") |>
  group_by(posteam, PlayType) |> summarize(count = n())

gt(passAndRunAttempts |> pivot_wider(names_from = posteam, values_from = count))
  
```

```{r}

passAndRunYds = nflData |> filter(PlayType == "Run" | PlayType == "Pass") |>
  select(Yards.Gained, PlayType)
head(passAndRunYds)


```

## Bar Chart - Are teams running the ball less?

```{r}

runPlaysCounts = nflDataClean |> select(Season, GameID, PlayType, Yards.Gained) |>
  filter(PlayType == "Run") |> group_by(GameID, Season) |> 
  summarize(runsPerGame = n()) |>
  ungroup() |>
  group_by(Season) |>
  summarize(avgRunsPerGame = mean(runsPerGame))

ggplot(runPlaysCounts, aes(x = avgRunsPerGame)) + geom_bar()


```

## Side By Side Bar Chart- Run Plays vs Pass Per Season

## Histogram: Rush Yards Per Game

```{r}
rushYdsPerGame = nflDataClean |> select(Season, GameID, PlayType, Yards.Gained) |>
  filter(PlayType == "Run") |> group_by(GameID) |> 
  summarize(ydsPerGame = sum(Yards.Gained))

g = ggplot(rushYdsPerGame, aes(x = ydsPerGame))

head(rushYdsPerGame)

g + geom_histogram()

```

## Scatter - Pass Yds vs Rush Yards Per Team Per Game:

```{r}

ydsYears = nflDataClean |> select(Season, GameID, PlayType, Yards.Gained, posteam) |>
  filter(PlayType == "Run" | PlayType == "Pass") |>
  drop_na(posteam) |>
  group_by(GameID, Season, posteam, PlayType) |>
  summarize(ydsPerPlay = sum(Yards.Gained))

summaryComp = ydsYears |> filter(PlayType == "Pass") |> select(Season, posteam, ydsPerPlay)

runs = ydsYears |> ungroup()|> filter(PlayType == "Run") |> select(ydsPerPlay) |> rename(rushYdsPerPlay = ydsPerPlay)


summaryComp$rushYdsPerPlay = runs$rushYdsPerPlay
summaryComp = summaryComp |> rename(passYdsPerPlay = ydsPerPlay) |> 
  mutate(teamYr = paste(Season, posteam, sep = "-"))


head(summaryComp)



g = ggplot(summaryComp, aes(x = rushYdsPerPlay, y = passYdsPerPlay))

g + geom_point(aes(color = posteam))


```

## Box Plot

## Faceted:

## Other Fun

## How would I extract final results of a game?

1.  Filter by team (home team == myTeam \| away team == myTeam)
2.  For each game (distinct(gameIds)
3.  Pick out "End of Game:
4.  Get Home Team Score and Away Team Score

```{r}
myTeam = "NYG"
yourTeam = "PHI"

## Get games involving both teams
myGames = as.tibble(nflDataClean[
  ((nflDataClean$HomeTeam == myTeam | nflDataClean$AwayTeam == myTeam) & (nflDataClean$HomeTeam == yourTeam | nflDataClean$AwayTeam == yourTeam)),])

#myGames = nflDataClean
#head(myGames)
#get vector of games
games = unique(myGames$GameID)

#print(mygames |> filter(PlayType == "End of Game"))


myGames = myGames |> filter(GameID %in% games) |> group_by(GameID) |> 
  drop_na(PosTeamScore, DefTeamScore) |>
  mutate(homeTeamScore = 
           ifelse(HomeTeam == posteam, PosTeamScore, DefTeamScore),
         awayTeamScore = 
           ifelse(AwayTeam == posteam, PosTeamScore, DefTeamScore)) |>
  mutate(homeTeamFinalScore = max(homeTeamScore),
         awayTeamFinalScore = max(awayTeamScore)) |>
  distinct(GameID, .keep_all = TRUE) |> select(Date,
                                       GameID,
                                       HomeTeam,
                                       homeTeamFinalScore,
                                       AwayTeam,
                                       awayTeamFinalScore
                                       ) |>
  mutate(winningTeam = factor(ifelse((homeTeamFinalScore > awayTeamFinalScore), as.character(HomeTeam), as.character(AwayTeam))))
  


gt(myGames |> ungroup()) |> tab_header("Matchup Results")

#resultsDF = tibble(colnames(c("homeTeam", "homeTeamScore", "awayTeam", "awayTeamScore")))
```

```{r}


wins = myGames |> mutate(
  homeTeamWon = ifelse(HomeTeam == winningTeam, 1, 0), 
  awayTeamWon = ifelse(AwayTeam == winningTeam, 1, 0)) |> group_by(HomeTeam)|> summarize(totalWins = sum(homeTeamWon + awayTeamWon))

g = ggplot(wins, aes(x = HomeTeam, y = totalWins))
g + geom_col(aes(fill = HomeTeam))
```

```{r}

## Time series - yds per game over time 

ydsYears = nflDataClean |> select(Date, Season, GameID, PlayType, Yards.Gained, posteam) |>
  filter(PlayType == "Run" | PlayType == "Pass") |>
  drop_na(posteam) |>
  group_by(Date, GameID, Season, posteam, PlayType) |>
  summarize(ydsPerPlay = sum(Yards.Gained))

summaryComp = ydsYears |> filter(PlayType == "Pass") |> select(Season, posteam, ydsPerPlay)

runs = ydsYears |> ungroup()|> filter(PlayType == "Run") |> select(ydsPerPlay) |> rename(rushYdsPerPlay = ydsPerPlay)


summaryComp$rushYdsPerPlay = runs$rushYdsPerPlay
summaryComp = summaryComp |> rename(passYdsPerPlay = ydsPerPlay) |> 
  mutate(teamYr = paste(Season, posteam, sep = "-"))


g = ggplot(summaryComp, aes(x = Date, y = passYdsPerPlay)) +geom_line(aes(group = posteam, color = posteam))

g + facet_wrap(posteam ~ .)

```
